---
title: "Fuel consumption comparison of automatic and manual transmission cars"
author: "Janos Brezniczky"
date: "27 December 2015"
output: pdf_document
---

```{r "data processing", echo = FALSE}
data(mtcars)
mtcars$am = factor(mtcars$am, levels = c(0, 1))
levels(mtcars$am) = c("automatic", "manual")
mtcars$vs = factor(mtcars$vs, levels = c(0, 1))
levels(mtcars$vs) = c("vertical", "straight")
```

## Executive Summary

The analysis of the given data (_mtcars_) did not suffice to prove a 
relationship between the transmission type (automatic/manual) and the miles per 
gallon (MPG). The final model suggests an increase in MPG of 0.276 (+- 2.75), 
which is equal to a random noise with a 84.2% chance.

To conclude with more confidence, more data, potentially about high MPG/automatic 
transmission and low MPG/manual transmission cars is needed.

## Exploratory Analysis

The analysis is based on the mtcars data set which is bundled with R. The R help
contains a description of the data set. It consists of 11 fields as seen in
figure 1. Except for the _V/S_ (_vs_) field interpretation was provided there, 
for that one a little investigation was necessary (see appendix).

For this analysis the _am_ variable is converted to a factor with levels
_"automatic"_ and _"manual"_, and _vs_ is similarly dealt with.

Having a look at the correlation matrix (fig. 2, 3) suggests that most variables 
are intercorrelated and thus to ultimately answer and quantify which transmission
type is more beneficial to the MPG of a car; confounder variables will need to 
be considered.

The MPG value seems to vary together with every other factor, given the 
correlation values, most weakly with the quarter mile time and the number of 
forward gears. The most significant correlations are experienced with
the weight, displacement and the number of cylinders variables.

## Impact of transmission type on consumption

Considering a simplistic model containing MPG as the outcome and only the 
transmission type as an explanatory variable is exhibited on figure 4. This 
chart both shows that the choice of automatic or manual transmission alone 
implies a big (around 5 MPG) change in fuel consumption and that there is little
or no data to estimate from about low MPG manual transmission cars.
The respective model (simple.model) accords with the first statement.

However, the aforementioned potential for confounders and the high correlation 
between MPG and other variables suggest to attempt adjusting for covariates.
The highest correlation and reasonable thinking (heavier cars are likely to 
consume more fuel, especially on inclines) suggest weight (wt) is a good 
choice.

## Augmented models

In the following, a series of nested models aim to refine the initial
conclusion. When adjusting for further variables, the significance of the AM
variable disappears and no further refinement recovers it.

1. Introducing weight into the model (fig. 5) removes the effect the 
transmission type formerly seemed to have. The p-value of AM increases to 0.988.
This suggests the variable is unnecessary, contrary to weight and the intercept.

2. The next plot suggests that there is a "U"-shaped trend in the residuals. One 
possible explanation is the presence of non-linearity. Note that from the recent
linear model one could have extrapolated as "heavy enough cars will have a 
negative MPG" due to coefficient signs, which is obvious nonsense. This is 
another way to reason for non-linearity in the response.
While the data set is likely to be insufficient for identifying the particular
non-linear relationship (investigating the domain could help), thinking of 
Taylor-polynomials suggests that the inclusion of higher order term(s) can 
easily be an adequate improvement.
Including the square of weight among the predictors seems to confirm this (fig. 
6), getting a coefficient of 1.3574 and a p-value well below 1%. The R-squared 
values also improve.

3. and 4. Looking for a pattern in the outliers (Fiat 128, etc.) shows (fig. 7)
that the three cars are very similar in horsepower. This is a convenient covariate 
candidate both numerically (it is near-continuous) and semantically (increasing 
hp and keeping all other aspect equal is a technological challenge to a 
designer, other factors are likely to degrade when having to compromise).
Including it and consequently its second power further improves the models 
giving much smaller residuals (cca. -2 to +4) than before (cca. -4 to +6).

## Justification of the model modifications

Analysis of variance (fig. 8) suggests that the nested improvements were 
correct, each iteration is acceptable on its own right.

According to the R-squared values, the explained part of the variation is above
85%, which, considering there may be known and unknown unknowns in explaining 
the MPG -- such as aerodynamic properties of the chassis, bearing and suspension 
quality; potentially a good achievement.

# Summary

With the given data, no reliable evidence of a relationship between 
transmission type and MPG has been found using even the most plausible 
refinement of the simplest straightforward model.
The model building procedure is confirmed by R-squared values, ANOVA tests and
p-values of variable-specific t-tests.

Any visible relationship of the type sought seems to be too weak compared to the 
prediction error, which -- if there is a relationship at all, is probably down 
to having a dataset too small in size, or too much focused on automatic 
transmission cars with low MPG and manual ones with a high MPG.

# Appendix: code and figures

### Figure 1: Data interpretation

Field|Interpetation|Field|Interpretation
-----|-------------|-----|--------------
mpg|Miles/(US) gallon|qsec|1/4 mile time
cyl|Number of cylinders|vs|V/S
disp|Displacement|am|Transmission (0 = automatic, 1 = manual)
hp|Gross horsepower|gear|Number of forward gears
drat|Rear axle ratio|carb|Number of careburetors
wt|Weight (1000 lbs)||

For vs, "0 means a V-engine, and 1 straight engine". Source:
http://stackoverflow.com/questions/18617174/r-mtcars-dataset-meaning-of-vs-variable

<!-- TODO: a ggplot2 alternative, if exists, is preferred (use alpha) -->
### Figure 2: Examining relationships - correlation matrix plot
```{r "figure2", fig.width=5, fig.height=3, echo = FALSE, fig.align = "center"}
#pairs(mtcars, gap = 0.2)
library(corrplot)
corrplot(cor(data.matrix(mtcars)), tl.cex = 0.5)
```

### Figure 3: Correlations with MPG
```{r "figure3", echo = FALSE}
cor.matrix = cor(data.matrix(mtcars))
# correlations with mpg
print(cor.matrix["mpg", ])
```

### Figure 4: Automatic vs. manual transmission fuel consumption

```{r "figure4", fig.width = 3, fig.height = 3, echo = FALSE}
plot(mtcars$am, mtcars$mpg, ylab = "MPG")
model.simple = lm(mpg ~ am, data = mtcars)
# print some summary
ss = summary(model.simple)
print(ss$coefficients)
cat("adjusted R-squared: ", ss$adj.r.squared, "\n")
```

### Figure 5: Automatic vs. manual fuel consumption modelled with vehicle weight
```{r "figure5plot", fig.height = 3, echo = FALSE,  fig.align="center"}
# add weight
model.nested = lm(mpg ~ am + wt, data = mtcars)

# summary(model.nested)
sn1 = summary(model.nested)
print(sn1$coefficients)
cat("adjusted R-squared: ", sn1$adj.r.squared, "\n")

# plot Residuals vs Fitted chart
plot.residuals = function(model, id.n = 0) {
  plot(model, which = c(1), main = "", cex.caption = 0.8, # cex.id = 0.4, 
       cex = 0.8, cex.oma.main = 0.7, sub.caption = "", cex.lab = 0.7, 
       cex.axis = 0.7, id.n = id.n)
}

plot.residuals(model.nested)
```

### Figure 6: Further nested models - non-linearity
```{r "figure 6", echo = FALSE, fig.height = 3}
model.nested.2 = update(model.nested, mpg ~ am + wt + I(wt ^ 2))
sn2 = summary(model.nested.2)
print(sn2$coefficients)
cat("adjusted R-squared: ", sn2$adj.r.squared, "\n")

# plot Residuals vs Fitted chart
plot.residuals(model.nested.2, id.n = 3)
# gives Merc 240D, Fiat 128, Toyota Corolla as the 3 biggest outliers
```

### Figure 7: Further nested models - adding horsepower
```{r "figure7", echo = FALSE, fig.height = 3}
print(mtcars[c("Fiat 128", "Merc 240D", "Toyota Corolla"), ])

model.nested.3 = lm(mpg ~ am + wt + I(wt ^ 2) + hp, data = mtcars)
cat("Nested model 3\n")
sn3 = summary(model.nested.3)
print(sn3$coefficients)
cat("adjusted R-squared: ", sn3$adj.r.squared, "\n")

model.nested.4 = lm(mpg ~ am + wt + I(wt ^ 2) + hp + I(hp ^ 2), data = mtcars)
cat("Nested model 4\n")
sn4 = summary(model.nested.4)
print(sn4$coefficients)
cat("adjusted R-squared: ", sn4$adj.r.squared, "\n")

plot.residuals(model.nested.4)
```

### Figure 8: Analysis of variance

```{r "figure 8", echo = FALSE}
anova(model.simple, model.nested, model.nested.2, model.nested.3, model.nested.4)
```
